---
title: "2.0-exploratory-data-analysis"
author: "Julia M.C."
date: "2026-04-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
# Loading the libraries and datasets
library(tidyverse)
library(lubridate)

crime <- read_csv("crime_cleaned.csv")
dispatch <- read_csv("dispatch_cleaned.csv")
```


# Severity Logic 
```{r}
unique(dispatch$initial_type)
unique(crime$crime_name1)
unique(crime$crime_name2)
```


```{r}
crime <- crime |>
  mutate(
    crime_rank = case_when(
      # RANK 3: Violent / Serious Crimes against Persons
      crime_name2 %in% c(
        "Forcible Rape", "Murder and Nonnegligent Manslaughter", "Kidnapping/Abduction", 
        "Aggravated Assault", "Robbery", "Human Trafficking, Commercial Sex Acts", 
        "Human Trafficking, Involuntary Servitude", "Negligent Manslaughter", 
        "Simple Assault", "Forcible Fondling", "Statutory Rape", "Incest", "Intimidation") ~ 3,

      # RANK 2: Property or Financial Related Crimes
      crime_name2 %in% c(
        "Motor Vehicle Theft", "Burglary/Breaking and Entering", "Arson", 
        "Weapon Law Violations", "Extortion/Blackmail", "Animal Cruelty",
        "Shoplifting", "All other Larceny", "Theft from Building", 
        "Theft From Motor Vehicle", "Theft of Motor Vehicle Parts or Accessories", 
        "Destruction/Damage/Vandalism of Property", "Embezzlement", "Stolen Property Offenses",
        "Hacking/Computer Invasion", "Identity Theft", "Wire Fraud", "Credit Card/Automatic Teller Machine Fraud") ~ 2,

      # RANK 1: Low / Minor non-violent / Drugs / Society
      crime_name2 %in% c(
        "Drug/Narcotic Violations", "Drug Equipment Violations", "Liquor Law Violations", 
        "Pornography/Obscene Material", "Prostitution", "Bribery", "Counterfeiting/Forgery", 
        "Impersonation", "Trespass of Real Property", "Disorderly Conduct", 
        "Curfew/Loitering/Vagrancy Violations", "Driving Under the Influence") ~ 1,

      # RANK 0: Non-Criminal / Service / Administrative
      crime_name2 %in% c(
        "Runaway", "Justifiable Homicide", "All Other Offenses") ~ 0,
      
      # Fallback logic based on broad categories
      crime_name1 == "Crime Against Person" ~ 3,
      crime_name1 == "Crime Against Property" ~ 2,
      crime_name1 == "Crime Against Society" ~ 1,
      TRUE ~ 0))
```

```{r}
crime |> 
  group_by(cr_number) |> 
  filter(n() > 1) |> 
  arrange(cr_number)
```

```{r}
# For each crime report keep only the highest severity crime (prevent duplicates)
crime_unique <- crime |>
  group_by(cr_number) |>
  slice_max(crime_rank, n = 1, with_ties = FALSE) |>
  ungroup()
```

```{r}
nrow(crime)
nrow(crime_unique)
```

```{r}
rank_dispatch_type <- function(type_col) {
  norm <- str_to_upper(str_squish(type_col))
  
  case_when(
    str_detect(norm, paste(c(
      "ASSAILANT", "SHOOTER", "RAPE", "SEXUAL ASSAULT", "SEX ASSAULT", "SEXASLT", 
      "ROBBERY JUST OCCURRED", "BURGLARY JUST OCCURRED", "CARJACKING", "CAR JACKING", 
      "SHOOTING", "STABBING", "STAB", "KIDNAPPING", "ABDUCTION", "HOSTAGE", 
      "CALLER IN DANGER", "PEDESTRIAN STRUCK", "PLANE DOWN", "HOUSE FIRE", 
      "TRAIN COLLISION", "DOMESTIC VIOLENCE", "DOMESTIC DISTURBANCE", "DOMVA", 
      "DOMVJA", "PURSUIT", "OFFICER NEEDS ASSIST", "WORKING CODE"), collapse = "|")) ~ 3,
    
    str_detect(norm, paste(c(
      "WEAPJ", "WEAPSUR", "WEAPONS", "FIREARMS", "BOMB", "CBRN", "EXPLODED", 
      "CONTAMINATION", "\\bDUI\\b", "INFLUENCE", "OVERDOSE", "SUICIDAL", "SUICIDE", 
      "VICIOUS", "ANIMALVJ", "HOLDUP", "DURESS", "PANIC", "ROBBERY", "ROBP", 
      "ASSAULT", "\\bASLT\\b", "ASLTA", "ASLTJA", "ASLTP", "WANTED PERSON", 
      "WANTED VEHICLE", "\\bCDS\\b", "CDSP", "ABUSEJ", "NEGLECTJ", "ABUSE", 
      "ABANDONMENT", "NEGLECT", "URGENT ASSIST", "ACCELERATOR STUCK", "HAZARDOUS",
      "MENTAL DISORDER", "MCOT"), collapse = "|")) ~ 2,
    
    str_detect(norm, paste(c(
      "THEFT", "LARCENY", "LACENY", "STLVEHT", "TRS", "SHOPLIFTING", "BURGLARY", 
      "BURGOS", "VANDALISM", "DAMAGE", "MISCHIEF", "VANDP", "FRAUD", "DECEPTION", 
      "FRAUDJ", "FRAUDT", "TRESPASSING", "UNWANTED", "TRESPJ", "TRESPOS", 
      "TRESP ON PATROL", "HARASSMENT", "STALKING", "THREATS", "HARASST", 
      "HARRASSOS", "THREATOS", "STALKOS", "PROSTITUTION", "LEWDNESS", 
      "INDECENCY", "HUNTING", "ILLEGAL", "ANIMAL ABUSE", "OCCURRED EARLIER", 
      "TRAFFIC VIOLATION", "TRAFFIC/TRANSPORTATION INCIDENT", "FIREWORKS", 
      "WEAPOS", "WANTED-TRS"), collapse = "|")) ~ 1,
    
    str_detect(norm, paste(c(
      "WELFARE", "ANI -", "ANIMAL RESCUE", "DISTURBANCE/NUISANCE", "NOISE", 
      "PARKING", "ADMIN", "DOCUMENT", "LOST", "FOUND", "PROPERTY", "STATION", 
      "MISCELLANEOUS", "MIS -", "E911", "CRYWOLF", "FOLLOW", "SUPPLEMENTAL", 
      "\\bASSIST\\b", "STANDBY", "RESCUE", "LOCK OUT", "ORDNANCE - FOUND", 
      "WARRANT SERVICE", "WS -", "GUARD", "ODJOB", "OFF DUTY", "S - SUSPICIOUS", 
      "COMMUNITY MEETING", "SIGNAL3", "TP -", "OAS", "HOMEDTN", "HOMEDTNT", 
      "MISSING", "RUNAWAY", "MP -", "TRANSPORT", "SUMMONS", "EVICTION", 
      "EXPARTE", "DETAIL", "DT -", "TD -", "TRAFFIC STOP", "TS -", "WR -", 
      "REPORT", "FARS", "VEHICLE ALARM", "ALARM", "10 SIGNAL", "SUSPICIOUS CIRC", 
      "DOMDOS", "DOMVOS", "DOMESTIC DISPUTE", "REFFERING"), collapse = "|")) ~ 0,
    TRUE ~ NA_real_)}
```

```{r}
combined_data <- dispatch |>
  mutate(
    initial_rank = rank_dispatch_type(initial_type)) |>
  left_join(
    crime_unique,
    by = c("crime_reports" = "cr_number"),
    suffix = c("_dispatch", "_crime")) |>
  mutate(
    police_district_number = coalesce(
      as.character(police_district_number_crime),
      as.character(police_district_number_dispatch)),
    year_month = year_month_dispatch,

    no_crime_report = if_else(is.na(crime_name1),
      "No Report", "Crime Report Generated"),

    initial_type_clean = str_to_upper(str_trim(initial_type)),
    crime_name1_clean = str_to_upper(str_trim(crime_name1)))
```


```{r}
# Create time metrics
combined_data <- combined_data |>
  mutate(
    start_time = ymd_hms(start_time),
    hour_of_day = hour(start_time),
    time_period = case_when(
      hour_of_day >= 6  & hour_of_day < 12 ~ "Morning",
      hour_of_day >= 12 & hour_of_day < 18 ~ "Afternoon",
      TRUE ~ "Night"))
```

```{r}
combined_data <- combined_data |>
  mutate(
    initial_norm = str_to_upper(str_squish(initial_type)),
    is_service_call = case_when(
      str_detect(initial_norm, 
                 "ADMIN|FOLLOW|SUPPLEMENTAL|ASSIST|DETAIL|TRANSPORT|STATION|WORKING CODE|WRITING REPORT|OFF DUTY|COMMUNITY MEETING|REFERRED|MESSAGES|DOCUMENT|LOST|FOUND|PETITION|SUMMONS|EVICTION|EXPARTE|PARKING|NOISE|DISTURBANCE|NUISANCE|SUSPICIOUS|WELFARE|CHECK|RESCUE|LOCK OUT|ALARM") ~ TRUE,
      TRUE ~ FALSE))
```

# Defining the perception gap
```{r}
combined_data <- combined_data |>
  mutate(
    severity_shift = case_when(
      no_crime_report == "No Report" ~ "No Report",
      is.na(initial_rank) | is.na(crime_rank) ~ "Unclassified",
      initial_rank == crime_rank ~ "Severity Match",
      initial_rank > crime_rank ~ "Severity Downgraded",
      initial_rank < crime_rank ~ "Severity Upgraded")) |>
  mutate(
    gap_no_crime = no_crime_report == "No Report",
    mismatch = severity_shift %in% c("Severity Downgraded", "Severity Upgraded")) |>
  mutate(
gap_type = case_when(
  is_service_call & no_crime_report == "No Report" ~ "Service / Non-Criminal",
  no_crime_report == "No Report" ~ "No Report",
  severity_shift == "Severity Downgraded" ~ "Severity Downgraded",
  severity_shift == "Severity Upgraded" ~ "Severity Upgraded",
  severity_shift == "Severity Match" ~ "Match",
  TRUE ~ "Unclassified"),

perception_gap_binary = case_when(
  severity_shift %in% c("Severity Downgraded", "Severity Upgraded", "No Report") ~ "Gap",
  severity_shift == "Severity Match" ~ "Accurate",
  TRUE ~ "Not Applicable"))

combined_data <- combined_data |>
  mutate(
    service_category = case_when(
      is_service_call & no_crime_report == "No Report" ~ "Service / Non-Criminal",
      is_service_call & no_crime_report == "Crime Report Generated" ~ "Service to Crime",
      TRUE ~ "Non-Service"))
prop.table(table(combined_data$severity_shift)) 
prop.table(table(combined_data$service_category))
```

# Aggregated Data (District-month)
```{r}
district_month_summary <- combined_data |>
  group_by(police_district_number, year_month) |>
  summarise(
  total_calls = n(),

  no_report_count = sum(severity_shift == "No Report"),
  no_report_rate = (no_report_count / total_calls) * 100,

  confirmed_crimes = sum(no_crime_report == "Crime Report Generated", na.rm = TRUE),

  mismatch_count = sum(severity_shift %in% c("Severity Downgraded", "Severity Upgraded"), na.rm = TRUE),
  mismatch_rate = if_else(
    confirmed_crimes > 0,
    (mismatch_count / confirmed_crimes) * 100, 0),

  upgraded_count = sum(severity_shift == "Severity Upgraded"),
  upgraded_rate = if_else(confirmed_crimes > 0, (upgraded_count / confirmed_crimes) * 100, 0),

  downgrade_count = sum(severity_shift == "Severity Downgraded"),
  downgrade_rate = if_else(confirmed_crimes > 0, (downgrade_count / confirmed_crimes) * 100, 0),

  total_gap_count = sum(severity_shift %in% c("No Report","Severity Upgraded","Severity Downgraded")),
  total_gap_rate = (total_gap_count / total_calls) * 100,
  .groups = "drop")
```

```{r}
final_table <- district_month_summary |>
  mutate(
    police_district_number = factor(police_district_number),
         year = year(year_month),
         month_num = month(year_month))

clean_data <- final_table |>
  filter(!is.na(no_report_rate), !is.na(no_report_count), !is.na(total_calls))

clean_combined <- combined_data |>
  filter(!is.na(severity_shift), !is.na(police_district_number), !is.na(initial_type)) |>
  mutate(police_district_number = factor(police_district_number),
        mismatch_binary = ifelse(mismatch == "Mismatch", 1, 0),
        gap_binary = ifelse(gap_no_crime == "No Crime Report", 1, 0),
        match_binary = ifelse(mismatch_binary == 1, 0, 1),
        report_binary = ifelse(gap_binary == 1, 0, 1))
```


```{r}
dim(final_table)
sum(is.na(final_table)) 
```


```{r}
head(final_table)
write_csv(final_table, "final_table3.0.csv")
write_csv(clean_data, "clean_data3.0.csv")
write_csv(clean_combined, "clean_combined3.0.csv")
```


# Visualizations
```{r}
ggplot(combined_data, aes(x = severity_shift, fill = severity_shift)) +
  geom_bar() +
  scale_fill_manual(values = c("No Report" = "#AB4832","Severity Downgraded" = "#CF9A1F", "Severity Match" = "#498C61", "Severity Upgraded" = "#577590", "Unclassified" = "#754CB5")) +
  labs(title = "Distribution of Severity Classification", x = "Severity Outcome", y = "Count", caption = "Source: Crime and Police Dispatched Incidents from DataMontgomery") +
  theme_minimal()
```
The majority of calls result in no crime report being generated. A much smaller number of cases result in confirmed crimes, with most of those being upgraded in severity. This indicates that the perception gap is not primarily driven by misclassification, but by the high volume of calls that do not lead to a crime report.


```{r}
combined_data |>
  filter(no_crime_report == "Crime Report Generated") |>
  ggplot(aes(x = severity_shift, fill = severity_shift)) +
  geom_bar() +
  scale_fill_manual(values = c("Severity Match" = "#498C61", "Severity Downgraded" = "#CF9A1F","Severity Upgraded" = "#577590", "Unclassified" = "#754CB5")) +
  labs(title = "Severity Classification (Confirmed Crimes Only)", x = "Severity Outcome", y = "Count", caption = "Source: Crime and Police Dispatched Incidents from DataMontgomery") +
  theme_minimal()
```
When focusing only on confirmed crimes, Severity Upgraded cases appear much more frequently than Severity Downgraded cases. This suggests that many incidents recorded by dispatch were later associated with more severe crime classifications in official police reports. Unclassified cases represent only a very small portion of the data.


```{r}
final_table |>
  filter(!is.na(police_district_number)) |>
  ggplot(aes(x = police_district_number, y = total_gap_rate)) +
  geom_boxplot(fill = "#90DBF4", color = "#023E8A", alpha = 0.8) +
  labs(title = "Perception Gap Distribution by District", x = "District", y = "Perception Gap Rate (%)", caption = "Source: Crime and Police Dispatched Incidents from DataMontgomery") +
  theme_minimal()
```
Perception gap rates remain consistently high across all police districts, with most districts showing median rates above 96%. Districts 1 and 2 have slightly higher gap rates, while Districts 5 and 6 tend to show somewhat lower levels overall. Several high outliers appear across districts. 

**The perception gap measure includes both classification mismatches and calls that did not result in a confirmed crime report, including many service-related or non-criminal incidents.**

```{r}
final_table |>
  filter(!is.na(police_district_number)) |>
  ggplot(aes(x = police_district_number, y = mismatch_rate)) +
  geom_boxplot(fill = "#C9B947", color = "#023E8A", alpha = 0.8) +
  labs(title = "Mismatch Rate by District", x = "District", y = "Mismatch Rate (%)", caption = "Source: Crime and Police Dispatched Incidents from DataMontgomery") +
  theme_minimal()
```
Mismatch rates vary across police districts. District 2 shows the highest typical mismatch rates, while Districts 1, 4, 5, and 6 tend to have somewhat lower and more similar distributions.These results suggest that classification mismatches occur throughout all districts, but some areas experience them more frequently than others.

```{r}
clean_data |>
  group_by(police_district_number) |>
  summarise(total_calls_sum = sum(total_calls)) |>
  ggplot(aes(x = police_district_number, y = total_calls_sum)) +
  geom_col(fill = "#637B87") +
  labs(title = "Total Calls by District", x = "District", y = "Total Calls", caption = "Source: Crime and Police Dispatched Incidents from DataMontgomery") +
  theme_minimal()
```
Total call volume varies across police districts. Districts 3 and 4 recorded the highest number of calls, while District 5 had the lowest overall call volume. These differences suggest that police demand is not evenly distributed across the county, with some districts consistently handling a larger share of dispatch activity than others.


```{r}
clean_data |>
  filter(!is.na(police_district_number)) |>
  ggplot(aes(x = police_district_number, y = total_calls)) +
  geom_boxplot(fill = "#9b2226", color = "black", alpha = 0.7) + 
  labs(title = "Monthly Call Volume Stability by District", subtitle = "Showing the distribution of calls per month for each district", x = "Police District", y = "Total Calls per Month", caption = "Source: Crime and Police Dispatched Incidents from DataMontgomery") +
  theme_minimal()
```
Monthly call volume patterns differ across police districts. Districts 3 and 4 show the highest  monthly call volumes, while District 5 consistently records the lowest levels of activity. 

```{r}
combined_data |>
  filter(perception_gap_binary == "Gap") |>
  ggplot(aes(x = gap_type, fill = gap_type)) +
  geom_bar() +
  scale_fill_manual(values = c("No Report" = "#AB4832", "Severity Downgraded" = "#4D9E51","Service / Non-Criminal" = "#E0C655", "Severity Upgraded" = "#6F3A78")) +
  labs(title = "Composition of the Perception Gap", x = "Gap Type", y = "Count", caption = "Source: Crime and Police Dispatched Incidents from DataMontgomery") +
  theme_minimal()
```
The perception gap is primarily driven by calls that do not result in confirmed crime reports. A large portion of these cases fall into the Service / Non-Criminal category, suggesting that many dispatch calls involve welfare checks, disturbances, administrative matters, or other non-criminal situations rather than confirmed criminal activity. 

