---
title: "The_Perception_Gap_Analysis"
author: "Julia M.C."
date: "2026-04-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
library(tidyverse)
library(lubridate)
library(pscl)

clean_data <- read_csv("clean_data3.0.csv") 
clean_combined <- read_csv("clean_combined3.0.csv")
```


# Hypotheses

## Null Hypothesis
Mean perception gap rates are equal across all police districts.

## Alternative Hypothesis 
At least one police district's mean perception gap rate is different from the others.

# Anova (District Differences)
```{r}
anova_model <- aov(total_gap_rate ~ factor(police_district_number), clean_data)

summary(anova_model)
```
- The results indicate statistically significant differences in perception gap rates across districts (we reject the null hypothesis).

## Tukey 
```{r}
TukeyHSD(anova_model)
```

- District 1 is significantly different from 4, 5, and 6.

- District 2 is also significantly different from 4, 5, and 6.

- District 3 is significantly different from 5 and 6.

District 2 vs. District 5 (diff = -0.64, p < 0.001): This is the largest recorded difference. District 5 has an average perception gap rate approximately 0.64 points lower than District 2.

District 1 vs. District 5 (diff = -0.48, p < 0.001): District 1 is significantly higher than the lower tier districts, specifically 4, 5, and 6.


Recommendations: 
Districts 1 and 2 have among the highest perception gap rates, so they may be useful starting points for further review.

Model Districts 5 and 6: These have the lowest gap rates. Investigation should be done to see what these districts are doing differently that could be applied to Districts 1 and 2.


# Linear Regression
```{r}
lm_fixed <- lm(total_gap_rate ~ factor(police_district_number) + factor(year), data = clean_data)
summary(lm_fixed)
plot(lm_fixed)
```
- About 24% of the variation in total gap rate is explained by district and year.

Most year coefficients are negative and significant compared to 2017, except 2023, which is not significant. This suggests that, regardless of the district, the total_gap_rate has been trending downward compared to 2017.

Districts 4, 5, and 6 still have significantly lower gap rates than District 1. This confirms what we saw in the ANOVA.

*Linearity and Variance (Residuals vs Fitted & Scale Location)*
The Residuals vs Fitted plot shows the red line is relatively flat, which is good for linearity.

However, the Scale Location plot shows a slight upward trend in the red line. This suggests that  predictions might be slightly less reliable at higher gap rates.

*Normality (Q-Q Residuals)*
The residuals curve away from the dashed line at both ends. This means the data has "heavy tails" (outliers).

Point 391 is a major outlier. It has a very large negative residual, meaning the model predicted a much higher gap rate than what actually occurred for that specific observation.

Residual diagnostics indicated slight heteroscedasticity and a few outliers, which are expected in real world public safety data. However, the diagnostic plots showed no severe violations that would invalidate the baseline regression approach.

# Logistic Regression

```{r}
logit_no_report <- glm(gap_no_crime ~ is_service_call + initial_rank + factor(police_district_number) + time_period, family = binomial, data = clean_combined)

summary(logit_no_report)

# Odds ratios
exp(coef(logit_no_report))
pR2(logit_no_report)
```
This model predicts whether a dispatch call ended without a matching crime report.

- Service / non-criminal calls were approximately 4.8 times more likely to result in a no report outcome compared to non-service calls.
- Higher severity calls were less likely to end without a crime report, suggesting that more serious incidents were more likely to generate official crime reports.
- Nighttime calls were associated with higher odds of no report outcomes, while morning calls were associated with slightly lower odds.
- District affects were also statistically significant, indicating that the probability of no report outcomes varied across police districts.

```{r}
logit_mismatch <- glm(mismatch ~ is_service_call + initial_rank + factor(police_district_number) + time_period, family = binomial, data = clean_combined)
summary(logit_mismatch)

# Odds ratios
exp(coef(logit_mismatch))
```
This model predicts whether the dispatch classification differed from the final crime classification.

- Service / non-criminal calls were less likely to be classification mismatches. This suggests that the perception gap is driven more by calls that never become formal crime reports than by classification mismatches.
- Higher severity was associated with a small increase in mismatch probability.
- Nighttime calls were associated with lower mismatch probabilities, while morning calls showed slightly higher mismatch odds.
- Several police districts showed significantly different mismatch probabilities compared to District 1.


**The results from both logistic regression models suggest that the perception gap is influenced more by the large number of dispatch calls that do not become formal crime reports than by differences in severity classification.**