---
title: "The_Perception_Gap_Analysis"
author: "Julia M.C."
date: "2026-04-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
library(tidyverse)
library(lubridate)
library(pscl)
library(caret)
library(pROC)

clean_data <- read_csv("clean_data3.0.csv") 
clean_combined <- read_csv("clean_combined3.0.csv")
```


# Hypotheses

## Null Hypothesis
Mean perception gap rates are equal across all police districts.

## Alternative Hypothesis 
At least one police district's mean perception gap rate is different from the others.

# Anova (District Differences)
```{r}
anova_model <- aov(total_gap_rate ~ factor(police_district_number), clean_data)
summary(anova_model)
```
- The results indicate statistically significant differences in perception gap rates across districts (we reject the null hypothesis).

# Tukey 
```{r}
TukeyHSD(anova_model)
```

District 2 vs. District 5 (diff = -0.64, p < 0.001): This is the largest recorded difference. District 5 has an average perception gap rate approximately 0.64 points lower than District 2.

District 1 vs. District 5 (diff = -0.48, p < 0.001): District 1 is significantly higher than the lower tier districts, specifically 4, 5, and 6.

Recommendations: 
Districts 1 and 2 have among the highest perception gap rates, so they may be useful starting points for further review.

Model Districts 5 and 6: These have the lowest gap rates. Investigation should be done to see what these districts are doing differently that could be applied to Districts 1 and 2.




# Multiple Linear Regression
```{r}
lm_fixed <- lm(total_gap_rate ~ factor(police_district_number) + factor(year),  data = clean_data)
summary(lm_fixed)
plot(lm_fixed)
```
- About 24% of the variation in total gap rate is explained by district and year.

Most year coefficients are negative and significant compared to 2017, except 2023, which is not significant. This suggests that, regardless of the district, the total_gap_rate has been trending downward compared to 2017.

Districts 4, 5, and 6 still have significantly lower gap rates than District 1. This confirms what we saw in the ANOVA.

*Linearity and Variance (Residuals vs Fitted & Scale Location)*
The Residuals vs Fitted plot shows the red line is relatively flat, which is good for linearity.

However, the Scale Location plot shows a slight upward trend in the red line. This suggests that  predictions might be slightly less reliable at higher gap rates.

*Normality (Q-Q Residuals)*
The residuals curve away from the dashed line at both ends. This means the data has "heavy tails" (outliers).

Point 391 is a major outlier. It has a very large negative residual, meaning the model predicted a much higher gap rate than what actually occurred for that specific observation.

Residual diagnostics indicated slight heteroscedasticity and a few outliers, which are expected in real world public safety data. However, the diagnostic plots showed no severe violations that would invalidate the baseline regression approach.

